---
title: "COVID death proportions by race and education"
author: "Kiran Bhattacharyya"
date: "2/15/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## COVID-19 death proportions by race and education

Here we will do some analysis on data released by CDC about how the total deaths due to COVID-19 differ between people of different racial categories and educational attainment. This data was collected between Jan 1, 2020 and Feb 1, 2021. It is open and can be downloaded from INSERT URL HERE.

First, let's load in the libraries and take a look at the data.

```{r get_data}
library(tidyverse)
library(ggplot2)

# read in covid data
covid_data <- read_csv('AH_Provisional_COVID-19_Deaths_by_Race_and_Educational_Attainment.csv')

# change column names so they don't have messy characters
names(covid_data) <- str_replace_all(names(covid_data), c(" " = "." , "-" = "" ))
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
